Towards a miniKanren with fair search strategies

### 2019-05-24 Paper/paper.tex
TOO mant corrections, paper is looking good

\usepackage{listings}
\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*,defrel,==,fresh,disj,conj,disj+},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true,
  literate={-inf}{{$^\infty$}}{1} {/fair}{{$_{fair}$}}{3}
  		   {→}{{$\rightarrow$}}{1}
  		   {×}{{$\times$}}{1}
  		   {==}{{$\equiv$}}{1}
  		   {conde}{{\textcolor[rgb]{0,.3,.7}{\textbf{cond$^e$}}}}{4}
           {appendo}{{append$^o$}}{6}
  		   {repeato}{{repeat$^o$}}{6}
  		   {disj2}{{disj$_2$}}{4}
  		   {conj2}{{conj$_2$}}{4}
  		   {g0}{{g$_0$}}{2}
  		   {g1}{{g$_1$}}{2}
  		   {g2}{{g$_2$}}{2}
  		   {g3}{{g$_3$}}{2}
  		   {gl}{{g$_l$}}{2}
  		   {gr}{{g$_r$}}{2}
  		   {DFSi }{{DFS$_{i}$}}{3}
  		   {DFSbi}{{DFS$_{bi}$}}{3}
  		   {DFSf }{{DFS$_{f}$}}{3}
}
%% scheme-list .

@@

\newcommand{\DFSi }[0]{DFS$_{i}$}
\newcommand{\DFSf }[0]{DFS$_{f}$}
\newcommand{\DFSbi}[0]{DFS$_{bi}$}
\newcommand{\BFS}[0]{BFS}
\newcommand{\BFSser}[0]{BFS}
\newcommand{\BFSuni}[0]{our BFS}
\newcommand{\BFSimp}[0]{our improved BFS}

@@

\NOTE[LKC]{
We actually have two new BFS implementations. One is modified from DFSf with 
two simple steps. It has been described briefly at the beginning of the BFS 
section but was not given a name. The other is the one described with two large 
figures and was given the name ``our improved BFS''. Both of them runs faster 
than the current BFS.

I didn't name the first one because (1) it is so simple and (2) I considered 
it merely a stepstone to our improved BFS. Now I decided to also give the first 
BFS implementation a name. This is because I think having a nameless impl is 
confusing.

Currently, the nameless implementation is given the name ``our BFS''. And I 
have changed the paper accordingly.

BUT, I am thinking about going further --- removing ``our improved BFS'' from 
this paper. The first reason is that our BFS is actually more efficient than 
our improved BFS in most benchmarks (see Table 1), despite its theoretical 
space extravagence. The second reason is that DFSi, DFSf, and our BFS look nice 
when sitting together.}

We describe fairness levels in disjunction and conjunction
implementations.  Specifically, a disjunction implementation can be
fair, almost-fair, or unfair. And a conjunction implementation can
be fair or unfair.  We compare the fairness level of four search
strategies: the standard miniKanren interleaving depth-first search,
the balanced interleaving depth-first search, the fair depth-first
search, and the standard breadth-first search.
The two non-standard depth-first searches are new. And we present two new, more 
efficient and shorter implementation of the standard breadth-first search. 
Using 
quantitative evaluation, we argue that the two new depth-first searches are 
competitive alternatives to the standard one, 
and that our breadth-first search implementations are more efficient than 
the current one.

\NOTE[LKC]{``simpler'' -> ``shorter'' (more objective)}

@@

\NOTE[DPF]{...  Mostly the point is to also include a reference to
the two dimitris and to the first OCamren (OCanren), whatever. It is up to you
in this paragraph.  ...}
\NOTE[LKC]{I have fixed the references to TRS and TRS2.}
\NOTE[LKC]{I have cited OCanren in this paragraph.}

\NOTE[DPF]{I have decided to agree with you that we should not use 
BFSser, so I am going to remove all occurrences.  
Also, we should say, ``Our implementation of BFS''
and then say, shortened to ``Our BFS'' and then you can use it in the table,
so one column should be Our BFS and the other should be BFS.}
\NOTE[DPF]{Occurrences of BFS (without Our) mean Seres et al. (We need only say
that once, or very occasionally}

\NOTE[DPF]{Is the spelling of reverso truly written without the last e?
Shouldn't it be reverseo with the appropriate raising of o.}
\NOTE[LKC]{Yes, it is reverso instead of reverseo in Dmitris' paper.}

@@

Next, consider the following disjunction of invoking \repeato{} with four 
different letters.

@@

\conde{} intuitively relates its clauses with logical \texttt{or}. And thus an 
unsuspicious beginner would expect each letter to contribute equally to the 
result, as follows.

@@

The \conde{} in TRS2, however, generates a less expected result.

@@

\DFSi{} is not always the best choice. For instance, it might be less 
desirable for little miniKanreners -- understanding implementation details and 
fiddling with clause order is not their first priority. 
There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves all
purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is generating simple and shallow programs,
the clauses for constructors had better be at the top of the
\conde{} expression.
When performing proof searches for complicated programs, the clauses for 
eliminators had better be at the top of the \conde{} expression. With \DFSi, 
these two uses cannot be efficient at the same time. In fact, to make one use 
efficient, the other one must be more sluggish.

@@

@@

Now we elaborate fairness by running more queries about \repeato{}. We never 
use \texttt{run*} in this paper because fairness is more interesting when we 
have an unbounded number of answers. However, it is perfectly fine to use 
\texttt{run*} with any search strategies.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the query result. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if the \conde{} would have
contained more clauses.

@@

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly 
more general
than the one in \citet{seres1999algebra}. Their definition is only
for binary disjunction. We generalize it to a multi-arity one.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer list. Search strategies with unfair \conj{} 
(e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.
Although some strategies have a different level of fairness in \disj{}, they 
have the same behavior when there is no call to a relational definition in 
\conde{} clauses (including this case).

@@

\NOTE[DPF]{Was this supposed to be a note?
``We never use \texttt{run*} in this paper because fairness is only interesting 
when we have an unbounded number of answers. However, it is perfectly fine to 
use run* with any search strategies.''}
\NOTE[LKC]{I initially wrote it down as a note. I might get it wrong, but I 
thought you suggested me to move it into the paper.}

@@

\NOTE[DPF]{Would the answer be the same if the list had been this ((a) (b))
or did you mean structurally the same?}
\NOTE[LKC]{I mean for all answer in the answer list, the answer's inner lists 
are pairwise \texttt{equal?}. It is impossible to see a \texttt{((a) (b))} in 
the answer list.}

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to this paper. TRS2,
chapter 10 and the appendix, ``Connecting the wires'', 
provides a comprehensive description of the 
miniKanren 
implementation but limited to unification constraints ($\equiv$).
Fig.~\ref{DFSi-0} and Fig.~\ref{DFSi-1} show parts that are later compared 
with other search 
strategies. We follow some conventions to name variables: \texttt{s}s name 
states; \texttt{g}s (possibly with subscript) name goals; 
variables ending with $^\infty$ name search spaces. Fig.~\ref{DFSi-0} shows the 
implementation of \disj. The 
first function, \disjtwo, implements binary disjunction. It applies the 
two disjunctive goals to the input state \texttt{s} and composes the two 
resulting search spaces with \texttt{append$^\infty$}. The following syntax 
definitions say \disj{} is right-associative. Fig.~\ref{DFSi-1} 
shows the implementation of \conj. The first function, \conjtwo{}, 
implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting search space. The helper function 
\texttt{append-map$^\infty$} applies its input goal to states 
in its input search spaces and composes the resulting search spaces. It reuses 
\texttt{append$^\infty$} for search space composition. The following syntax 
definitions say \conj{} is also right-associative.

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\disjtwo{} (Fig.~\ref{fDFS}). The new \disjtwo{} calls a new and 
fair version of \texttt{append$^\infty$}. \texttt{append$^\infty_{fair}$} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that we haven't swapped
\texttt{s$^\infty$} and \texttt{t$^\infty$}. The swapping 
happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s$^\infty$} and 
\texttt{t$^\infty$} are ended with a thunk, and that we have swapped them. In 
this case, we construct a new thunk. The new thunk swaps two spaces back in the
recursive call to \texttt{loop}. This is unnecessary for fairness. We do it to 
produce answers in a more natural order.

\NOTE[DPF]{When or where do you swap them back? This could be still clearer.}
\NOTE[LKC]{Improved.}

@@

This implementation is improvable in two ways. First, as mentioned in 
\autoref{sec:fairconj}, \BFS{} puts answers in bags and answers of the same 
cost are in the same bag. In the above implementation, however, 
it is not very clear how we manage cost information --- the \texttt{car}s of a 
search space have cost 0 (i.e., they are all in the same bag), and every thunk 
indicates an increment in cost. It is even more subtle that 
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects the cost information. Second, \texttt{append$^\infty_{fair}$} is 
extravagant in memory usage. It makes $O(n+m)$ new \texttt{cons} cells every 
time, where $n$ and $m$ are the ``length''s of input spaces. \DFSf{} also has 
the space extravagance.

@@

We make the cost information more clear by changing the type of search space, 
modifying related function definitions, and introducing a few more functions.

@@

\NOTE[DPF]{I am not in love with the two calls to null?. I can think of at least
two ways to fix it.  Pick one.}
\NOTE[LKC]{fixed}

@@

\NOTE[DPF]{The use of the references to Fig 7 twice tells me that something
has to change for the sake of clarity.}
\NOTE[LKC]{I guess you are talking about Table 1 becasue there is no Fig 7. I 
have change the first reference.}

@@

\NOTE[DPF]{If we forget to use {} after \reverso, the o ends up too near
the next word.  I fixed reverso on the second line of section 7, but there 
may be others.}
\NOTE[LKC]{I have checked all uses of similar macros by text search.}

@@

\NOTE[LKC]{I ref the next section instead of the table}

@@

\NOTE[DPF]{There is still some clumsiness using our improved DFS, since
at some point, we may want to make this clearer, so have DFS not be fonted
is a little weird.  I think both should always use ourDFS and 
macro expand it to <our improved DFS>.  Then changes will be much easier.}
\NOTE[LKC]{I guess you are talking about ``our improved BFS''. I have added 
relevant macros.}

\NOTE[LKC]{I rewrote the following paragraph and the table to include our BFS 
in additional to our improved BFS.}
\NOTE[LKC]{I also re-measured all running time. I used to run benchmarks with 
DrRacket, now I do it in command-line. Thus, the time measurements are smaller 
and more stable.}

@@

\NOTE[DPF]{We might find the reference Ocanren in the two Dmitri's paper
and say something here about it.}
\NOTE[LKC]{added}

>>>>

\usepackage{listings}
\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*defrel,==,fresh,disj,conj,disj+},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true,
  literate={-inf}{{$^\infty$}}{1} {/fair}{{$_{fair}$}}{3}
  		   {→}{{$\rightarrow$}}{1}
  		   {×}{{$\times$}}{1}
  		   {==}{{$\equiv$}}{1}
  		   {conde}{{\textcolor[rgb]{0,.3,.7}{\textbf{cond$^e$}}}}{4}
           {appendo}{{append$^o$}}{6}
  		   {repeato}{{repeat$^o$}}{6}
  		   {disj2}{{disj$_2$}}{4}
  		   {conj2}{{conj$_2$}}{4}
  		   {g0}{{g$_0$}}{2}
  		   {g1}{{g$_1$}}{2}
  		   {g2}{{g$_2$}}{2}
  		   {g3}{{g$_3$}}{2}
  		   {gl}{{g$_l$}}{2}
  		   {gr}{{g$_r$}}{2}
  		   {DFSi }{{DFS$_{i}$}}{3}
  		   {DFSbi}{{DFS$_{bi}$}}{3}
  		   {DFSf }{{DFS$_{f}$}}{3}
}
%% scheme-list .

@@

\newcommand{\DFSi }[0]{DFS$_{i}$}
\newcommand{\DFSf }[0]{DFS$_{f}$}
\newcommand{\DFSbi}[0]{DFS$_{bi}$}
\newcommand{\BFS}[0]{BFS}
\newcommand{\BFSser}[0]{BFS}
\newcommand{\BFSuni}[0]{our BFS}
\newcommand{\BFSimp}[0]{our improved BFS}
\newcommand{\runstar}{run$^*$}

@@

\NOTE[DPF]{Throughout the paper we have search spaces and search strategies.
I can see leaving in the  ``search strategies'', but let's try to
remove virtually all of the uses of ``search'' in front of spaces.}

\NOTE[LKC]{BUT, I am thinking about going further --- removing ``our improved'' from 
this paper. The first reason is that our BFS is actually more efficient than 
our improved BFS in most benchmarks (see Table 1), despite its theoretical 
space extravagence. The second reason is that DFSi, DFSf, and our BFS look nice 
when sitting together.}

\NOTE[DPF]{Does this mean that the comment about append goes away.
I really thought that was so clever.  I thought that the way you pointed
out the trivial changes was very clear, so that did not bother me at
all.}

We describe fairness levels in disjunction and conjunction
implementations.  Specifically, a disjunction implementation can be
fair, almost-fair, or unfair. And a conjunction implementation can be
fair or unfair.  We compare the fairness level of four search
strategies: the standard miniKanren interleaving depth-first search,
the balanced interleaving depth-first search, the fair depth-first
search, and the standard breadth-first search.  The two non-standard
depth-first searches are new. And we present two new, more efficient
and shorter implementations of the standard breadth-first search.
Using quantitative evaluation, we argue that the two new depth-first
searches are competitive alternatives to the standard one, and that
our breadth-first search implementations are more efficient than the
current one.

@@

\NOTE[DPF]{their BFS (or shortened to BFS). We need only say this once if we go with ``our'', which I don't like.
I like ``the new'' or ``the improved'' and ``the current'' }

@@

\noindent Next, consider the following disjunction of invoking \repeato{} with four 
different letters.

@@

\noindent \conde{} intuitively relates its clauses with logical \texttt{or}. And thus an 
unsuspicious beginner would expect each letter to contribute equally to the 
result, as follows.

@@

\noindent The \conde{} in TRS2, however, generates a less expected result.

@@

\DFSi{} is not always the best choice. For instance, it might be less 
desirable for novice miniKanrenuser---understanding implementation details and 
fiddling with clause order is not their first priority. 
There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves all
purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is generating simple and shallow programs,
the clauses for constructors had better be at the top of the
\conde{} expression.
When performing proof searches for complicated programs, the clauses for 
eliminators had better be at the top of the \conde{} expression. With \DFSi, 
these two uses cannot be efficient at the same time. In fact, to make one use 
efficient, the other one must be more sluggish.

@@

\NOTE[DPF]{The citetseres1999algebra must be not broken:
            You can either rewrite the sentence or
            use a tilde called a tie that means don't break the line.
            Sometimes using tilde does not work out elegant enough,
            so the first idea works best.}

@@

Now we elaborate fairness by running more queries about \repeato{}. We never 
use \runstar{} in this paper because fairness is more interesting when we 
have an unbounded number of answers. It is perfectly fine, however, to use 
\runstar{} with any search strategies.

@@

Given the following program, it is natural to expect lists of each
letter to constitute $1/4$ in the query result. \DFSi, TRS2's search
strategy, however, results in many more lists of \texttt{a}s than
lists of other letters. And some letters (e.g. \texttt{c} and
\texttt{d}) are rarely seen. The more clauses the worse the situation.

@@

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly 
more general
than the one in \citet{seres1999algebra}, which is only
for binary disjunction. We generalize it to a multi-arity one.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer list. Search strategies with unfair \conj{}:
\DFSi, \DFSbi, and \DFSf, however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. Here again, as the
number of clauses grows, the situation worsens.

Although some strategies have a different level of fairness in \disj{}, they 
have the same behavior when there is no call to a relational definition in 
\conde{} clauses (including this case).

@@

@@

\NOTE[DPF]{Can you express exactly how all costs are determined?  I see that
for this simple program, your notion of cost makes sense, but how does it
generalize to all costs.}
\NOTE[LKC]{I mean for all answer in the answer list, the answer's inner lists 
are pairwise \texttt{equal?}. It is impossible to see a \texttt{((a) (b))} in 
the answer list.}

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to this paper. TRS2,
chapter 10 and the appendix, ``Connecting the wires'', 
provide a comprehensive description of the 
miniKanren 
implementation but limited to unification constraints ($\equiv$).
Fig.~\ref{DFSi-0} and Fig.~\ref{DFSi-1} show parts that are later compared 
with other search 
strategies. We follow some conventions to name variables: \texttt{s}s name 
states; \texttt{g}s (possibly with subscript) name goals; 
variables ending with $^\infty$ name (search) spaces. Fig.~\ref{DFSi-0} shows the 
implementation of \disj. The 
first function, \disjtwo, implements binary disjunction. It applies the 
two disjunctive goals to the input state \texttt{s} and composes the two 
resulting search spaces with \texttt{append$^\infty$}. The following syntax 
definitions say \disj{} is right-associative. Fig.~\ref{DFSi-1} 
shows the implementation of \conj. The first function, \conjtwo{}, 
implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting search space. The helper function 
\texttt{append-map$^\infty$} applies its input goal to states 
in its input search spaces and composes the resulting search spaces. It reuses 
\texttt{append$^\infty$} for search space composition. The following syntax 
definitions say \conj{} is also right-associative.

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\disjtwo{} (Fig.~\ref{fDFS}). The new \disjtwo{} calls a new and 
fair version of \texttt{append$^\infty$}. \texttt{append$^\infty_{fair}$} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that we haven't swapped
\texttt{s$^\infty$} and \texttt{t$^\infty$}. The swapping 
happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s$^\infty$} and 
\texttt{t$^\infty$} are ended with a thunk, and that we have swapped them. In 
this case, we construct a new thunk. The new thunk swaps back the two spaces in the
recursive call to \texttt{loop}. This is unnecessary for fairness---we do it to 
produce answers in a more readable order.

@@

\NOTE[DPF]{Why are the two texttt's macros, again?  Now, it hardly matters,
but it is unfriendly to your other co-authors to have us reading long
expressions when a newcommand would suffice.  I read the paper with a pen
marking it up, then I read the paper on the .tex screen.  That's each time
I read it.}

\NOTE[DPF]{You know that ``bag'' is a technical terms.  It means that there
is a structure where each element of the structure has an associated number, 
which tells you have many occurrences of that element are in the list.
For example, the list (a b a c a d c) could be represented like this:
(a$^3$ b$^1$ c$^2$ d$^1$).  If that is not what you mean, you might change the name
to bucket.}

This implementation can be improved in two ways. First, as mentioned in 
\autoref{sec:fairconj}, \BFS{} puts answers in bags and answers of the same 
cost are in the same bag. In the above implementation, however, 
it is not obvious how we manage cost information---the \texttt{car}s of a 
search space have cost 0 (i.e., they are all in the same bag), and every thunk 
indicates an increment in cost. It is even more subtle that 
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects the cost information. Second, \texttt{append$^\infty_{fair}$} is 
extravagant in memory usage. It makes $O(n+m)$ new \texttt{cons} cells every 
time, where $n$ and $m$ are the ``length''s of input spaces. \DFSf{} also has 
the space extravagance.

@@

\NOTE[DPF]{Font Space like in the code comments.}

We make the cost information more clear by changing the Space type, 
modifying related function definitions, and introducing a few more functions.

@@

@@

\NOTE[DPF]{ks --> sk and kf --> fk and you don't need to say f and s
mean success continuation and failure continuation.}

@@

@@

@@

\NOTE[DPF]{We don't have disequality constraints, so no absento.  You might just
say that.}

@@
